\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{hyperref}

\newcommand{\Ctr}{\mathcal{C}}
\newcommand{\Host}{\mathcal{H}}
\newcommand{\RHost}[1]{\mathcal{R}_{#1}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Compile}{\mathsf{Compile}}
\newcommand{\Judge}{\mathsf{Judge}}
\newcommand{\Hist}{\mathsf{Hist}}
\newcommand{\View}{\mathsf{View}}
\newcommand{\Worldline}{\mathsf{Worldline}}
\newcommand{\Braid}{\mathsf{Braid}}
\newcommand{\WarpState}{\mathsf{WarpState}}
\newcommand{\Witness}{\mathsf{Witness}}
\newcommand{\Receipt}{\mathsf{Receipt}}
\newcommand{\Real}{\mathsf{Real}}
\newcommand{\TTD}{\ensuremath{\mathsf{TTD}}}
\newcommand{\Echo}{\ensuremath{\mathsf{Echo}}}
\newcommand{\GitWarp}{\ensuremath{\mathsf{GitWarp}}}
\newcommand{\Fuse}{\ensuremath{\mathsf{Fuse}}}

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{observation}{Observation}
\newtheorem{principle}{Principle}
\newtheorem{proposition}{Proposition}

\title{Continuum Categories\\[6pt]
\large Contract Categories, Realization Profunctors, and Witness Squares\\[3pt]
\normalsize 2026-04-09}
\author{}
\date{}

\begin{document}
\maketitle

\section{Context}

These notes record a minimal categorical story for the Continuum stack.
The target is not a maximal formalism. The target is the smallest
defensible account that explains:
\begin{enumerate}[leftmargin=2em]
\item what Wesley compiles,
\item what the host runtimes consume,
\item why the right semantic shape is profunctorial,
\item how witness output upgrades generation into conformance.
\end{enumerate}

The intended runtime family is the present Continuum surface:
\begin{itemize}[leftmargin=2em]
\item \Echo{} as a Continuum participant,
\item \GitWarp{} as a Continuum participant,
\item \TTD{} as the host-neutral debugger protocol surface,
\item \Fuse{}-style materializations as host adapters over worldlines, strands,
  or braids.
\end{itemize}

Continuum is the protocol for exchanging witnessed causal history. It is not a
runtime hierarchy, and it does not make \Echo{} the owner of \GitWarp{}'s
participant role.

\section{Design Claim}

\begin{principle}[Continuum split]
The authored contract surface, the host runtime presentation surface, and the
conformance witness surface are distinct layers. Treating any one of them as
the others produces shadow authority and contract drift.
\end{principle}

\begin{observation}[Wesley is not one profunctor]
Wesley should not be modeled as a single profunctor. Wesley is better modeled
as a compiler that, for each host runtime, produces a realization profunctor,
a canonical compiled presentation, and a local witness that the presentation
realizes the authored contract.
\end{observation}

\section{The Contract Category}

\begin{definition}[Authored contract category]
Let $\Ctr$ be the category of authored Continuum contracts.

Objects of $\Ctr$ are admitted semantic contract families: protocol schemas,
observer envelopes, receipt families, coordinate families, rewrite
declarations, footprint declarations, and related shared noun families.

A morphism
\[
\alpha : c \to c'
\]
in $\Ctr$ is a \emph{refinement embedding} from a weaker contract $c$ into a
stronger contract $c'$. Concretely, $c'$ preserves the obligations of $c$ and
adds further structure, fields, laws, or distinctions.
\end{definition}

This direction is deliberate. If $c'$ refines $c$, then any runtime surface
that realizes $c'$ should also realize $c$ after forgetting the extra
structure. The variance of the realization profunctor below captures exactly
that rule.

\begin{observation}[SDL is syntax, not the category]
GraphQL SDL is a concrete authored syntax for presenting objects of $\Ctr$.
It is not itself the semantic category. The category lives at the level of
admitted contract meaning, not raw source files.
\end{observation}

\section{Host Presentation Categories}

\begin{definition}[Host presentation category]
For each host $H \in \Host$, let $\RHost{H}$ be the category of runtime
presentations for that host.

Objects of $\RHost{H}$ are host-facing generated or stabilized surfaces:
TypeScript types, Rust types, manifests, codecs, protocol envelopes,
artifact bundles, adapter stubs, and other host-native realizations.

A morphism
\[
\beta : r \to r'
\]
in $\RHost{H}$ is a lawful host-side adaptation that preserves the semantic
meaning required by the contract family under discussion.
\end{definition}

Examples of hosts include:
\[
H \in \{\Echo,\ \GitWarp,\ \TTD,\ \Fuse\}.
\]

\begin{observation}[Host categories do not own contract truth]
The host presentation categories carry realized surfaces, not authored
authority. Runtime convenience, storage shape, and protocol framing belong in
the host category only after the authored contract boundary has been named.
\end{observation}

\section{Realization Profunctors}

\begin{definition}[Realization profunctor]
For each host $H$, define a profunctor
\[
\Real_H : \Ctr^{op} \times \RHost{H} \to \Set
\]
where $\Real_H(c,r)$ is the set of witnesses that runtime presentation $r$
faithfully realizes contract $c$ for host $H$.
\end{definition}

The two variances carry the core Continuum rules:
\begin{itemize}[leftmargin=2em]
\item \textbf{Contravariant in the contract.} If
  $\alpha : c \to c'$ is a refinement embedding, then a witness for the
  stronger contract $c'$ induces a witness for the weaker contract $c$.
\item \textbf{Covariant in the runtime presentation.} If
  $\beta : r \to r'$ is a lawful host adaptation, then a witness for $r$
  transports to a witness for $r'$.
\end{itemize}

So for every $\alpha : c \to c'$ in $\Ctr$ and $\beta : r \to r'$ in
$\RHost{H}$, the profunctor yields a transport map
\[
\Real_H(\alpha,\beta) : \Real_H(c',r) \to \Real_H(c,r').
\]

\begin{observation}[Why the profunctor shape is the right one]
The contract side and the runtime side do not vary in the same direction.
Strengthening a contract moves opposite to realization transport, while lawful
runtime adaptation moves with it. That asymmetry is exactly what makes the
realization story profunctorial rather than functorial.
\end{observation}

\section{Canonical Compilation}

\begin{definition}[Host-indexed compile path]
For each host $H$, Wesley provides a contravariant compile path
\[
\Compile_H : \Ctr_{\mathrm{adm}}^{op} \to \RHost{H}
\]
defined on the admissible contract families carried by that compile path.
\end{definition}

Objectwise, $\Compile_H(c)$ is the canonical compiled presentation of contract
$c$ for host $H$.

On a refinement embedding $\alpha : c \to c'$, the arrow
\[
\Compile_H(\alpha) : \Compile_H(c') \to \Compile_H(c)
\]
is the host-side forgetting or adaptation map from the richer compiled surface
to the weaker one.

\begin{definition}[Witness assignment]
For each admissible contract $c$, Wesley emits a local witness
\[
\mathsf{wit}_H(c) \in \Real_H\bigl(c,\Compile_H(c)\bigr).
\]
\end{definition}

\begin{proposition}[Witness-square law]
For each refinement embedding $\alpha : c \to c'$ in the admissible contract
surface, the compiled surfaces and witnesses satisfy
\[
\Real_H\!\bigl(\alpha,\Compile_H(\alpha)\bigr)\bigl(\mathsf{wit}_H(c')\bigr)
= \mathsf{wit}_H(c).
\]
\end{proposition}

This is the core conformance condition. It says that the witness for the
stronger compiled contract, transported across the contract refinement and the
runtime adaptation, agrees with the witness for the weaker compiled contract.

\begin{observation}[Why witness output matters]
Generation without the witness-square law is only code emission. The witness
surface is what upgrades the compile path into a conformance claim.
\end{observation}

\section{Publication Boundaries}

The categorical story also clarifies Wesley's publication-boundary role.

\begin{principle}[Publication-boundary rule]
For an admitted shared noun family, the repo must name:
\begin{enumerate}[leftmargin=2em]
\item the authored home carrying the contract object in $\Ctr$,
\item the host-indexed compile path $\Compile_H$ that consumes it,
\item the stable generated surface $\Compile_H(c)$ expected by consumers,
\item the local witness $\mathsf{wit}_H(c)$ or equivalent conformance output.
\end{enumerate}
\end{principle}

If any of these are missing, the system has target-state intent but not a
fully admitted contract lane.

\section{Observer Geometry and Optics}

The present account also separates three ideas that should not be collapsed.

\begin{enumerate}[leftmargin=2em]
\item \textbf{Observer Geometry} studies the geometry of projection:
  observers, apertures, degeneracies, translation costs, mediator paths, and
  observer-relative logical preservation.

\item \textbf{WARP optics} studies the full lawful rewrite envelope around a
  projection. In the current working notation, a WARP optic is a tuple
  \[
  \Omega = (\pi,\mathcal{F},\rho,\omega,\sigma)
  \]
  consisting of projection, focus boundary, local rewrite, witness, and lawful
  reintegration.

\item \textbf{Wesley} is the contract compiler that turns authored shared
  surfaces into host-indexed runtime presentations and witness squares. Wesley
  does not own the runtime semantics of observers or optics; it preserves their
  authored contract boundary and realizes them coherently across hosts.
\end{enumerate}

\begin{observation}[Projection versus full optic]
Observer Geometry is primarily about the geometry of $\pi$.
Optics studies the larger object in which $\pi$ lives.
Wesley compiles authored contracts that may name either projection-only
surfaces or richer optic-bearing surfaces, but it should not be identified
with the optic itself.
\end{observation}

\section{Host-Indexed Optic Interpretations}

The profunctor account becomes especially useful once authored contracts carry
rewrite and footprint structure.

An authored optic-bearing contract may then be interpreted across hosts by
different realization contexts:
\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
Host or realization context & Interpreted role \\
\midrule
\Echo{} & execute Echo-local rewrites \\
\GitWarp{} & admit and materialize causal patches \\
\TTD{} & expose debugger playback and explanation surfaces \\
\Fuse{} & materialize editable filesystem projections \\
footprint analysis & derive read/write/delete scope \\
witness extraction & derive local reversibility residue \\
\bottomrule
\end{tabular}
\end{center}

The important point is not that these are identical implementations. The
important point is that they are host-indexed realizations of one authored
contract family rather than handwritten shadows with approximately matching
names.

\section{Judgment Surfaces Are Downstream}

Wesley's judgment-bridge role can also be expressed categorically, but it is a
downstream stage.

For each host $H$, one may define a category $\mathcal{J}_H$ of
operator-facing judgment surfaces and a functor
\[
\Judge_H : \RHost{H} \to \mathcal{J}_H
\]
that turns substrate-facing realized surfaces into operator-facing statuses,
risk classes, confidence adjustments, explanations, or gate results.

\begin{observation}[Judgment does not replace realization]
Judgment surfaces do not substitute for contract realization witnesses. They
consume realized runtime facts. They do not retroactively become the fact
layer.
\end{observation}

\section{What This Story Does Not Yet Claim}

\begin{itemize}[leftmargin=2em]
\item It does not claim a complete bicategory or equipment of Continuum
  contracts, observers, and optics.
\item It does not yet define full optic laws for authored rewrites.
\item It does not identify a canonical enrichment for observer distance,
  aperture, or budgeted translation cost.
\item It does not prove that every practical Wesley compile path already
  satisfies the witness-square law; that remains a repo-by-repo burden of
  evidence.
\end{itemize}

\section{Future Lift}

If the stack matures further, the natural next lift is toward a double
categorical picture:
\begin{itemize}[leftmargin=2em]
\item objects: contract domains or admitted noun families,
\item vertical arrows: structure-preserving refinements,
\item horizontal arrows: realization relations, translators, or observer
  relations,
\item squares: witness-bearing coherence between refinement and realization.
\end{itemize}

But the present note does not need that extra machinery to be useful. The
realization profunctor plus witness-square law already explains why Wesley's
role is stronger than plain code generation and narrower than substrate
ownership.

\section{One-Sentence Summary}

Continuum is best modeled by a contract category, host-indexed runtime
presentation categories, realization profunctors between them, and local
witness squares showing that Wesley's compiled surfaces really realize the
authored contracts they claim to carry.

\end{document}
